---
title: "Climate-Related Hazards Damage SHELDUS data"
format: html
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

## Data Source:

SPATIAL HAZARD EVENTS AND LOSSES DATABASE FOR THE UNITED STATES (SHELDUS) Version 23.0

https://sheldus.asu.edu/

Downloaded: July 3, 2025

## Key Variables:

-   cropdmg: Direct damage to crops in nominal U.S. dollars.

-   cropdmg_adj: Direct damage to crops in adjusted 2023 U.S. dollars.

-   CropDmgPerCapita: Crop damage per capita, adjusted to the most recent SHELDUS inflation base year.

-   propertydmg: Direct damage to property in nominal U.S. dollars (no decimals).

-   propertydmg_adj: Direct damage to property in adjusted 2023 U.S. dollars (two decimal places).

-   PropertyDmgPerCapita: Property damage per capita, adjusted to the most recent SHELDUS inflation base year.

-   duration_days: Duration of the event in days (integer).

## Hazard Categories (18 total):

1.  Avalanche
2.  Coastal
3.  Drought
4.  Earthquake
5.  Flood
6.  Fog
7.  Hail
8.  Heat
9.  Hurricane/Tropical Storm
10. Landslide
11. Lightning
12. Severe Thunderstorm
13. Tornado
14. Tsunami/Seiche
15. Volcano
16. Wildfire
17. Wind
18. Winter Weather Load Libraries and Data

```{r}
library(tidyverse)
library(here)
library(gt)
library(scales)

# Load and clean SHELDUS data
sheldus <- read_csv(here("data", "direct_loss_aggregated_output_5008.csv"))
names(sheldus) <- tolower(names(sheldus))

sheldus_clean <- sheldus %>%
  rename(
    state = statename,
    cropdmg_adj = `cropdmg(adj 2023)`,
    propertydmg_adj = `propertydmg(adj 2023)`
  ) %>%
  select(state, year, hazard, cropdmg = cropdmg_adj, propertydmg = propertydmg_adj)
```

## Summary Statistics

### Annual Damage Summary

```{r}
summary_by_year <- sheldus_clean %>%
  group_by(year) %>%
  summarize(
    total_property_dmg = sum(propertydmg, na.rm = TRUE),
    total_crop_dmg = sum(cropdmg, na.rm = TRUE),
    .groups = "drop"
  )

summary_by_year |>
  mutate(
    total_property_dmg = total_property_dmg / 1e6,
    total_crop_dmg = total_crop_dmg / 1e6
  ) |>
  gt() |>
  tab_header(title = "Annual Direct Damage Summary (Adjusted to 2023 Dollars)") |>
  fmt_currency(columns = starts_with("total"), currency = "USD") |>
  cols_label(
    total_property_dmg = "Total Property Damage (million)",
    total_crop_dmg = "Total Crop Damage (million)"
  )
```

### Damage by State

```{r}
summary_by_state <- sheldus_clean %>%
  group_by(state) %>%
  summarize(
    total_property_dmg = sum(propertydmg, na.rm = TRUE),
    total_crop_dmg = sum(cropdmg, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_property_dmg))

summary_by_state |>
  mutate(
    total_property_dmg = total_property_dmg / 1e6,
    total_crop_dmg = total_crop_dmg / 1e6
  ) |>
  gt() |>
  tab_header(title = "Total Damage by State (Adjusted to 2023 Dollars)") |>
  fmt_currency(columns = starts_with("total"), currency = "USD") |>
  cols_label(
    total_property_dmg = "Total Property Damage (million)",
    total_crop_dmg = "Total Crop Damage (million)"
  )
```

### Damage by Hazard Type

```{r}
summary_by_hazard <- sheldus_clean %>%
  group_by(hazard) %>%
  summarize(
    total_property_dmg = sum(propertydmg, na.rm = TRUE),
    total_crop_dmg = sum(cropdmg, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_property_dmg))

summary_by_hazard |>
  mutate(
    total_property_dmg = total_property_dmg / 1e6,
    total_crop_dmg = total_crop_dmg / 1e6
  ) |>
  gt() |>
  tab_header(title = "Total Damage by Hazard Type (Adjusted to 2023 Dollars)") |>
  fmt_currency(columns = starts_with("total"), currency = "USD") |>
  cols_label(
    total_property_dmg = "Total Property Damage (million)",
    total_crop_dmg = "Total Crop Damage (million)"
  )
```

## Climate-Related Hazard Focus

Damage Trends: Most Climate-Related Hazards, including "Drought", "Flood", "Heat", "Hurricane/Tropical Storm", "Wildfire", "Coastal"

```{r}
most_climate_related_hazards <- c(
  "Drought",
  "Flood",
  "Heat",
  "Hurricane/Tropical Storm",
  "Wildfire",
  "Coastal"
)

plot_top_hazards <- sheldus_clean %>%
  filter(hazard %in% most_climate_related_hazards) %>%
  mutate(total_dmg = propertydmg + cropdmg) %>%
  mutate(total_dmg = total_dmg / 1e6) %>%
  group_by(year, hazard) %>%
  summarize(total_dmg = sum(total_dmg, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = year, y = total_dmg)) +
  geom_line(color = "steelblue") +
  facet_wrap(~ hazard, scales = "free_y") +
  scale_y_continuous(labels = label_dollar()) +
  labs(
    title = "Most Climate-Related Hazards: Annual Damage Trends",
    subtitle = "Total direct damage (property + crop), adjusted to 2023 dollars",
    x = "Year",
    y = "Total Damage (2023 USD, in million)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )

plot_top_hazards
```

## Data ready for merge

```{r}
sheldus_formerge <- sheldus |>
  select(
    statename,
    year,
    hazard,
    cropdmg,
    "cropdmg(adj 2023)",
    cropdmgpercapita,
    propertydmg,
    "propertydmg(adj 2023)",
    propertydmgpercapita,
    duration_days
  ) |>
  rename(
    cropdmg_adj = "cropdmg(adj 2023)",
    propertydmg_adj = "propertydmg(adj 2023)",
    state = statename
  )


summary(sheldus_formerge)
```

## Descriptive Statistics

```{r}
library(psych)

describe(sheldus_formerge |> select(-state, -year, -hazard))

saveRDS(sheldus_formerge, here::here("data", "sheldus.rds"))
write.csv(sheldus_formerge, here::here("data", "sheldus.csv"))


library(haven)
write_dta(sheldus_formerge, here::here("data", "sheldus.dta"))

```

sheldus.RData for R sheldus.csv for Excel sheldus.dta for Stata
